---
title: "Supporting Information 3"
author: "Daniel Y.M. Nakamura, Ward C. Wheeler, Taran Grant"
date: "2025-01-25"
output: html_document
---

# PRE-AMBLE

```{r setup, include=FALSE}
# Set the working directory 
knitr::opts_knit$set(root.dir = "/Users/labanfibios/Desktop/Doutorado/Project/B2_TEvsMOL/GitHub/")

# Load libraries
library(ape)
library(cowplot)
library(dplyr)
library(dunn.test)
library(ggpubr)
library(ggridges)
library(glmmTMB)
library(gridExtra)
library(phytools)
library(RNODE)
library(tidyverse)
library(TreeDist)
library(viridis)
```

# Q1. Are topologies impacted by the addition of phenomic evidence?

Dataset with MOL-TE topological distances were calculated in script 1.

```{r}
# Load data
df <- read.csv("Dataset_R_q123_06-03-25.csv")

# Number of datasets impacted by the addition of phenomic evidence
sum(df$MP_MOLvsTE_NormalizedRF > 0) / nrow(df) # MP MOL vs MP TE
sum(df$ML_MOLvsTEASC_NormalizedRF > 0) / nrow(df) # ML MOL vs ML TE ASC
sum(df$ML_MOLvsTEnoASC_NormalizedRF > 0) / nrow(df) # ML MOL vs ML TE noASC
```

The phenomic evidence impacted topologies in 90.3% (47/52) of the datasets in MP and ML (with ASC) and 88.5% (46/52) in ML (without ASC).

## Q1.1 Data exploration

```{r}
# 1.a MPmol_MPte
# No. shared clades
mean(df$MP_MOLvsTE_nSharedClades)
sd(df$MP_MOLvsTE_nSharedClades)
range(df$MP_MOLvsTE_nSharedClades)
# No. unique MOL clades
mean(df$MP_nUniqueCladesMOL)
sd(df$MP_nUniqueCladesMOL)
range(df$MP_nUniqueCladesMOL)
# No. unique TE clades
mean(df$MP_nUniqueCladesTE)
sd(df$MP_nUniqueCladesTE)
range(df$MP_nUniqueCladesTE)
# RF
mean(df$MP_MOLvsTE_NormalizedRF)
sd(df$MP_MOLvsTE_NormalizedRF)
range(df$MP_MOLvsTE_NormalizedRF)
# CID
mean(df$MP_MOLvsTE_NormalizedCID)
sd(df$MP_MOLvsTE_NormalizedCID)
range(df$MP_MOLvsTE_NormalizedCID)
# SPR
mean(as.numeric(df$MP_MOLvsTE_NormalizedSPR))
sd(as.numeric(df$MP_MOLvsTE_NormalizedSPR))
range(as.numeric(df$MP_MOLvsTE_NormalizedSPR))
# Resolution MOL
mean(as.numeric(df$MP_nPolytomiesMOL))
sd(as.numeric(df$MP_nPolytomiesMOL))
range(as.numeric(df$MP_nPolytomiesMOL))
# Resolution TE
mean(as.numeric(df$MP_nPolytomiesTE))
sd(as.numeric(df$MP_nPolytomiesTE))
range(as.numeric(df$MP_nPolytomiesTE))

# 1.b MLmol_MLteAsc
# No. shared clades
mean(df$ML_MOLvsTEASC_nSharedClades)
sd(df$ML_MOLvsTEASC_nSharedClades)
range(df$ML_MOLvsTEASC_nSharedClades)
# No. unique MOL clades
mean(df$ML_MOLvsTEASC_nUniqueCladesMOL)
sd(df$ML_MOLvsTEASC_nUniqueCladesMOL)
range(df$ML_MOLvsTEASC_nUniqueCladesMOL)
# No. unique TE clades
mean(df$ML_MOLvsTEASC_nUniqueCladesTEASC)
sd(df$ML_MOLvsTEASC_nUniqueCladesTEASC)
range(df$ML_MOLvsTEASC_nUniqueCladesTEASC)
# RF
mean(df$ML_MOLvsTEASC_NormalizedRF)
sd(df$ML_MOLvsTEASC_NormalizedRF)
range(df$ML_MOLvsTEASC_NormalizedRF)
# CID
mean(df$ML_MOLvsTEASC_NormalizedCID)
sd(df$ML_MOLvsTEASC_NormalizedCID)
range(df$ML_MOLvsTEASC_NormalizedCID)
# SPR
mean(as.numeric(df$ML_MOLvsTEASC_NormalizedSPR))
sd(as.numeric(df$ML_MOLvsTEASC_NormalizedSPR))
range(as.numeric(df$ML_MOLvsTEASC_NormalizedSPR))

# 1.c MLmol_MLteNoAsc
# No. shared clades
mean(df$ML_MOLvsTEnoASC_nSharedClades)
sd(df$ML_MOLvsTEnoASC_nSharedClades)
range(df$ML_MOLvsTEnoASC_nSharedClades)
# No. unique MOL clades
mean(df$ML_MOLvsTEnoASC_nUniqueCladesMOL)
sd(df$ML_MOLvsTEnoASC_nUniqueCladesMOL)
range(df$ML_MOLvsTEnoASC_nUniqueCladesMOL)
# No. unique TE clades
mean(df$ML_MOLvsTEnoASC_nUniqueCladesTEnoASC)
sd(df$ML_MOLvsTEnoASC_nUniqueCladesTEnoASC)
range(df$ML_MOLvsTEnoASC_nUniqueCladesTEnoASC)
# RF
mean(df$ML_MOLvsTEnoASC_NormalizedRF)
sd(df$ML_MOLvsTEnoASC_NormalizedRF)
range(df$ML_MOLvsTEnoASC_NormalizedRF)
# CID
mean(df$ML_MOLvsTEnoASC_NormalizedCID)
sd(df$ML_MOLvsTEnoASC_NormalizedCID)
range(df$ML_MOLvsTEnoASC_NormalizedCID)
# SPR
mean(as.numeric(df$ML_MOLvsTEnoASC_NormalizedSPR))
sd(as.numeric(df$ML_MOLvsTEnoASC_NormalizedSPR))
range(as.numeric(df$ML_MOLvsTEnoASC_NormalizedSPR))
```

## Q1.2 Statistical tests

In MP and ML (with ASC), there are 47 datasets in which MOL-TE incongruences are not equal to zero and 5 datasets in which they are equal to zero. In ML (without ASC), there are 46 datasets in which MOL-TE incongruences are not equal to zero and 6 datasets in which they are equal to zero. To test if MOL-TE incongruences are significantly different from zero, we can check if the distribution of observed values are significantly different from zero. 

For each comparison and for each topological distance metric, we tested for normality. Because data are not normal according to Shapiro-Wilk tests (P < 0.05 in all cases), we tested if the median of each topological distance metric is different from zero using a non-parametric Wilcoxon Signed-Ranked test. First, the V statistic is computed by summing the signed ranks of the differences between the observed values and the hypothesized value (i.e. zero). The ranks are assigned based on the absolute values of the differences, and the sign is determined by whether the original value is positive or negative. Second, given that the distribution of V values is normally distributed for a large sample size due to the Central Limit Theorem, P-values can be calculated by finding the area under the normal curve corresponding to the calculated Z-score.

```{r}
# Load data
df <- read.csv("Dataset_R_q123_06-03-25.csv")

# 1a. MPmol vs MPte
# RF
shapiro.test(as.numeric(df$MP_MOLvsTE_NormalizedRF)) # non-parametric, p < 0.05
q1a_rf <- wilcox.test(as.numeric(df$MP_MOLvsTE_NormalizedRF), mu = 0, exact=F, conf.int=T)
q1a_rf
# CID
shapiro.test(as.numeric(df$MP_MOLvsTE_NormalizedCID))  # non-parametric, p < 0.05
q1a_cid <- wilcox.test(as.numeric(df$MP_MOLvsTE_NormalizedCID), mu = 0, exact=F, conf.int=T)
q1a_cid
# SPR
shapiro.test(as.numeric(df$MP_MOLvsTE_NormalizedSPR)) # non-parametric, p < 0.05
q1a_spr <- wilcox.test(as.numeric(df$MP_MOLvsTE_NormalizedSPR), mu = 0, exact=F, conf.int=T)
q1a_spr

# 1b. MLmol vs MLteAsc
#RF
shapiro.test(as.numeric(df$ML_MOLvsTEASC_NormalizedRF)) # non-parametric, p < 0.05
q1b_rf <- wilcox.test(as.numeric(df$ML_MOLvsTEASC_NormalizedRF), mu = 0, exact=F, conf.int=T)
q1b_rf
# CID
shapiro.test(as.numeric(df$ML_MOLvsTEASC_NormalizedCID)) # non-parametric, p < 0.05
q1b_cid <- wilcox.test(as.numeric(df$ML_MOLvsTEASC_NormalizedCID), mu = 0, exact=F, conf.int=T)
q1b_cid
# SPR
shapiro.test(as.numeric(df$ML_MOLvsTEASC_NormalizedSPR)) # non-parametric, p < 0.05
q1b_spr <- wilcox.test(as.numeric(df$ML_MOLvsTEASC_NormalizedSPR), mu = 0, exact=F, conf.int=T)
q1b_spr

# 1c. MLmol vs MLteNoAsc
# RF
shapiro.test(as.numeric(df$ML_MOLvsTEnoASC_NormalizedRF)) # non-parametric, p < 0.05
q1c_rf <- wilcox.test(as.numeric(df$ML_MOLvsTEnoASC_NormalizedRF), mu = 0, exact=F, conf.int=T)
q1c_rf
# CID
shapiro.test(as.numeric(df$ML_MOLvsTEnoASC_NormalizedCID)) # non-parametric, p < 0.05
q1c_cid <- wilcox.test(as.numeric(df$ML_MOLvsTEnoASC_NormalizedCID), mu = 0, exact=F, conf.int=T)
q1c_cid
# SPR
shapiro.test(as.numeric(df$ML_MOLvsTEnoASC_NormalizedSPR)) # non-parametric, p < 0.05
q1c_spr <- wilcox.test(as.numeric(df$ML_MOLvsTEnoASC_NormalizedSPR), mu = 0, exact=F, conf.int=T)
q1c_spr
```
In all cases, the median of normalized RF, CID, and SPR values in MP and ML are significantly different from zero. Thus, MOL-TE incongruences different from zero indicate that the phenomic evidence tends to impact topologies in empirical studies.

## Q1.3 Figure

```{r}
# Load data
df <- read.csv("Dataset_R_q123_06-03-25.csv")

# Create a dataframe
df <- data.frame(
  RFdist = c(df$MP_MOLvsTE_NormalizedRF, df$ML_MOLvsTEASC_NormalizedRF, df$ML_MOLvsTEnoASC_NormalizedRF),  # Combine the vectors into a single column
  CID = c(df$MP_MOLvsTE_NormalizedCID, df$ML_MOLvsTEASC_NormalizedCID, df$ML_MOLvsTEnoASC_NormalizedCID),
  SPR = c(df$MP_MOLvsTE_NormalizedSPR, df$ML_MOLvsTEASC_NormalizedSPR, df$ML_MOLvsTEnoASC_NormalizedSPR),
  comparison = rep(c("MPmol_MPte", "MLmol_MLteAsc", "MLmol_MLteNoAsc"), each = length(df$MP_MOLvsTE_NormalizedRF))  # Repeat vector names
)
df$comparison <- factor(df$comparison, levels = c("MPmol_MPte", "MLmol_MLteAsc", "MLmol_MLteNoAsc"))
custom_palette <- c("MPmol_MPte" = "#339999",
                    "MLmol_MLteAsc" = "#6f77b4",
                    "MLmol_MLteNoAsc" = "#cccccc")
custom_labels <- c(
  "MPmol_MPte" = "MP-MOL vs MP-TE",
  "MLmol_MLteAsc" = "ML-MOL vs ML-TE (with ASC)",
  "MLmol_MLteNoAsc" = "ML-MOL vs ML-TE (without ASC)"
)
s1_rf = ggplot(df, aes(x=RFdist, y=comparison, fill=comparison))+
  geom_density_ridges() +
  scale_x_continuous(limits = c(0, NA)) +  # Set lower limit to 0, prevents extrapolation below 0
  theme_ridges()+
  labs(x="Normalized RF", y=NULL,tag="A")+
  scale_fill_manual(values = custom_palette, labels = custom_labels)+
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 1, alpha=.5) +
  theme(
    axis.text.y = element_blank(),  # Remove y-axis values
    legend.position = "none",     # Move the legend below the plot
    legend.direction = "horizontal", # Display legend items horizontally (side by side)
    legend.box = "horizontal",      # Make sure the legend is laid out horizontally
    legend.title = element_blank(),  # Optional: Remove legend title
    legend.text = element_text(size = 8),  # Reduce the size of legend text
    legend.key.size = unit(0.5, "cm"),    # Reduce the size of legend keys
    axis.title.x = element_text(hjust = 0.5)  # Center the x-axis title
  )
s1_cid = ggplot(df, aes(x=CID, y=comparison, fill=comparison))+
  geom_density_ridges() +
  scale_x_continuous(limits = c(0, NA)) +  # Set lower limit to 0, prevents extrapolation below 0
  theme_ridges()+
  labs(x="Normalized CID", y=NULL,tag="B")+
  scale_fill_manual(values = custom_palette, labels = custom_labels)+
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", size = 1, alpha=.5) +
  theme(
    axis.text.y = element_blank(),  # Remove y-axis values
    legend.position = "none",     # Move the legend below the plot
    legend.direction = "horizontal", # Display legend items horizontally (side by side)
    legend.box = "horizontal",      # Make sure the legend is laid out horizontally
    legend.title = element_blank(),  # Optional: Remove legend title
    legend.text = element_text(size = 8),  # Reduce the size of legend text
    legend.key.size = unit(0.5, "cm"),    # Reduce the size of legend keys
    axis.title.x = element_text(hjust = 0.5)  # Center the x-axis title
  )
s1_spr = ggplot(df, aes(x=SPR, y=comparison, fill=comparison))+
  geom_density_ridges() +
  scale_x_continuous(limits = c(0, NA)) +  # Set lower limit to 0, prevents extrapolation below 0
  theme_ridges()+
  labs(x="Normalized SPR", y=NULL,tag="C")+
  scale_fill_manual(values = custom_palette, labels = custom_labels)+
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", size = 1, alpha=.5) +
  theme(
    axis.text.y = element_blank(),  # Remove y-axis values
    legend.position = "none",     # Move the legend below the plot
    legend.direction = "horizontal", # Display legend items horizontally (side by side)
    legend.box = "horizontal",      # Make sure the legend is laid out horizontally
    legend.title = element_blank(),  # Optional: Remove legend title
    legend.text = element_text(size = 8),  # Reduce the size of legend text
    legend.key.size = unit(0.5, "cm"),    # Reduce the size of legend keys
    axis.title.x = element_text(hjust = 0.5)  # Center the x-axis title
  )
# Extract the legend as a separate ggplot object
legend <- ggplot(df, aes(x = comparison, y = CID, color = comparison)) +
  geom_point() +
  scale_color_manual(values = custom_palette, labels = custom_labels) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
  )+
  guides(color = guide_legend(override.aes = list(size = 2, alpha=.7)))  # Increase dot size in legend
legend_grob <- cowplot::get_legend(legend)

#tiff("Figures/figQ1_density.tiff", width = 6, height = 8, units = "in", res = 300)
grid.arrange(arrangeGrob(s1_rf + theme(plot.margin = margin(.01,1,1,.01)),
                         s1_cid + theme(plot.margin = margin(.01,1,1,.01)),
                         s1_spr + theme(plot.margin = margin(.01,1,1,.01)),
                         nrow=3, ncol=1),
             legend_grob,
             nrow=2,
             heights= c(20,1))
#dev.off()
```

# Q2.Are MOL-TE incongruences associated with the proportion of phenomic characters? 

## Q2.1 Spearman's Rank Correlations

To test if the proportion of phenomic characters is associated with the topological metrics of MOL-TE incongruences, we used a non-parametric Spearman's Rank Correlation test.

```{r, warning=F}
# Load data
df <- read.csv("Dataset_R_q123_06-03-25.csv")

# RF: MP MOL vs MP TE
cor.test(df$X.Morph.Total, df$MP_MOLvsTE_NormalizedRF, method = "spearman")
plot_mp_rf = ggplot(df, aes(x = X.Morph.Total, y = MP_MOLvsTE_NormalizedRF)) +
  geom_point() +  # scatter plot
  geom_smooth(method = "lm", formula = y ~ x, color = "#339999", se = FALSE) +  # regression line (lm) for the trend
  stat_cor(method = "spearman", label.x = 0.08, label.y = 0.6) +  # add Spearman correlation value
  theme_minimal()+
  labs(y = "Normalized RF\n", x = NULL)

# CID: MP MOL vs MP TE
cor.test(df$X.Morph.Total, df$MP_MOLvsTE_NormalizedCID, method = "spearman")
plot_mp_cid = ggplot(df, aes(x = X.Morph.Total, y = MP_MOLvsTE_NormalizedCID)) +
  geom_point() +  # scatter plot
  geom_smooth(method = "lm", formula = y ~ x, color = "#339999", se = FALSE) +  # regression line (lm) for the trend
  stat_cor(method = "spearman", label.x = 0.08, label.y = 0.6) +  # add Spearman correlation value
  theme_minimal()+
  labs(y = "Normalized CID\n", x = NULL)

# SPR: MP MOL vs MP TE
cor.test(df$X.Morph.Total, df$MP_MOLvsTE_NormalizedSPR, method = "spearman")
plot_mp_spr = ggplot(df, aes(x = X.Morph.Total, y = MP_MOLvsTE_NormalizedSPR)) +
  geom_point() +  # scatter plot
  geom_smooth(method = "lm", formula = y ~ x, color = "#339999", se = FALSE) +  # regression line (lm) for the trend
  stat_cor(method = "spearman", label.x = 0.08, label.y = 0.33) +  # add Spearman correlation value
  theme_minimal()+
  labs(y = "Normalized SPR\n", x = "\n% Phenomic characters")
```

```{r, warning=F}
# RF: ML MOL vs ML TE (ASC)
cor.test(df$X.Morph.Total, df$ML_MOLvsTEASC_NormalizedRF, method = "spearman")
plot_mlasc_rf = ggplot(df, aes(x = X.Morph.Total, y = ML_MOLvsTEASC_NormalizedRF)) +
  geom_point() +  # scatter plot
  geom_smooth(method = "lm", formula = y ~ x, color = "#6f77b4", se = FALSE) +  # regression line (lm) for the trend
  stat_cor(method = "spearman", label.x = 0.09, label.y = 0.73) +  # add Spearman correlation value
  theme_minimal()+
  labs(x = NULL, y = NULL) # Remove x and y axis titles

# CID: ML MOL vs ML TE (ASC)
cor.test(df$X.Morph.Total, df$ML_MOLvsTEASC_NormalizedCID, method = "spearman")
plot_mlasc_cid = ggplot(df, aes(x = X.Morph.Total, y = ML_MOLvsTEASC_NormalizedCID)) +
  geom_point() +  # scatter plot
  geom_smooth(method = "lm", formula = y ~ x, color = "#6f77b4", se = FALSE) +  # regression line (lm) for the trend
  stat_cor(method = "spearman", label.x = 0.09, label.y = 0.515) +  # add Spearman correlation value
  theme_minimal()+
  labs(x = NULL, y = NULL)  # Remove x and y axis titles
  
# SPR: ML MOL vs ML TE (ASC)
cor.test(df$X.Morph.Total, df$ML_MOLvsTEASC_NormalizedSPR, method = "spearman")
plot_mlasc_spr = ggplot(df, aes(x = X.Morph.Total, y = ML_MOLvsTEASC_NormalizedSPR)) +
  geom_point() +  # scatter plot
  geom_smooth(method = "lm", formula = y ~ x, color = "#6f77b4", se = FALSE) +  # regression line (lm) for the trend
  stat_cor(method = "spearman", label.x = 0.09, label.y = 0.334) +  # add Spearman correlation value
  theme_minimal()+
  labs(y = NULL, x = "\n% Phenomic characters")
```

```{r, warning=F}
# RF: ML MOL vs ML TE (noASC)
cor.test(df$X.Morph.Total, df$ML_MOLvsTEnoASC_NormalizedRF, method = "spearman")
plot_mlno_rf = ggplot(df, aes(x = X.Morph.Total, y = ML_MOLvsTEnoASC_NormalizedRF)) +
  geom_point() +  # scatter plot
  geom_smooth(method = "lm", formula = y ~ x, color = "#cccccc", se = FALSE) +  # regression line (lm) for the trend
  stat_cor(method = "spearman", label.x = 0.09, label.y = 0.73) +  # add Spearman correlation value
  theme_minimal()+
  labs(x = NULL, y = NULL)  # Remove x and y axis titles

# CID: ML MOL vs ML TE (noASC)
cor.test(df$X.Morph.Total, df$ML_MOLvsTEnoASC_NormalizedCID, method = "spearman")
plot_mlno_cid = ggplot(df, aes(x = X.Morph.Total, y = ML_MOLvsTEnoASC_NormalizedCID)) +
  geom_point() +  # scatter plot
  geom_smooth(method = "lm", formula = y ~ x, color = "#cccccc", se = FALSE) +  # regression line (lm) for the trend
  stat_cor(method = "spearman", label.x = 0.09, label.y = 0.525) +  # add Spearman correlation value
  theme_minimal()+
  labs(x = NULL, y = NULL)  # Remove x and y axis titles

# SPR: ML MOL vs ML TE (noASC)
cor.test(df$X.Morph.Total, df$ML_MOLvsTEnoASC_NormalizedSPR, method = "spearman")
plot_mlno_spr = ggplot(df, aes(x = X.Morph.Total, y = ML_MOLvsTEnoASC_NormalizedSPR)) +
  geom_point() +  # scatter plot
  geom_smooth(method = "lm", formula = y ~ x, color = "#cccccc", se = FALSE) +  # regression line (lm) for the trend
  stat_cor(method = "spearman", label.x = 0.085, label.y = 0.42) +  # add Spearman correlation value
  theme_minimal()+
  labs(y = NULL, x = "\n% Phenomic characters")
```

## Q2.2 Figure

```{r}
# Grid 3x3
#tiff("Figures/figQ2_proportion.tiff", width = 12, height = 10, units = "in", res = 300)
grid.arrange(arrangeGrob(plot_mp_rf + theme(plot.margin = margin(.01,1,1,.01)),
                         plot_mlasc_rf + theme(plot.margin = margin(.01,1,1,.01)),
                         plot_mlno_rf + theme(plot.margin = margin(.01,1,1,.01)),
                         plot_mp_cid + theme(plot.margin = margin(.01,1,1,.01)),
                         plot_mlasc_cid + theme(plot.margin = margin(.01,1,1,.01)),
                         plot_mlno_cid + theme(plot.margin = margin(.01,1,1,.01)),
                         plot_mp_spr + theme(plot.margin = margin(.01,1,1,.01)),
                         plot_mlasc_spr + theme(plot.margin = margin(.01,1,1,.01)),
                         plot_mlno_spr + theme(plot.margin = margin(.01,1,1,.01)),
                         nrow=3, ncol=3),
             legend_grob,
             nrow=2,
             heights= c(20,1))
#dev.off()
```


# Q3. Is the addition of phenomic evidence equally impactful on ML and MP topologies?

## Q3.1 MOL-TE incongruences ~ Optimality criteria

To test whether the MOL-TE incongruences are different between ML and MP, we first need to change the format of the dataframe.

```{r}
# Load data
df <- read.csv("Dataset_R_q123_06-03-25.csv")

# Create an empty data frame for RF
q3_rf <- data.frame(value = numeric(0), source = character(0))
# Combine the vectors of RF into the data frame
q3_rf <- rbind(q3_rf, data.frame(value = df$MP_MOLvsTE_NormalizedRF, source = rep("MP_MOLvsTE", length(df$MP_MOLvsTE_NormalizedRF))))
q3_rf <- rbind(q3_rf, data.frame(value = df$ML_MOLvsTEASC_NormalizedRF, source = rep("ML_MOLvsTEASC", length(df$ML_MOLvsTEASC_NormalizedRF))))
q3_rf <- rbind(q3_rf, data.frame(value = df$ML_MOLvsTEnoASC_NormalizedRF, source = rep("ML_MOLvsTEnoASC", length(df$ML_MOLvsTEnoASC_NormalizedRF))))

# Create an empty data frame for CID
q3_cid <- data.frame(value = numeric(0), source = character(0))
# Combine the vectors of CID into the data frame
q3_cid <- rbind(q3_cid, data.frame(value = df$MP_MOLvsTE_NormalizedCID, source = rep("MP_MOLvsTE", length(df$MP_MOLvsTE_NormalizedCID))))
q3_cid <- rbind(q3_cid, data.frame(value = df$ML_MOLvsTEASC_NormalizedCID, source = rep("ML_MOLvsTEASC", length(df$ML_MOLvsTEASC_NormalizedCID))))
q3_cid <- rbind(q3_cid, data.frame(value = df$ML_MOLvsTEnoASC_NormalizedCID, source = rep("ML_MOLvsTEnoASC", length(df$ML_MOLvsTEnoASC_NormalizedCID))))

# Create an empty data frame for SPR
q3_spr <- data.frame(value = numeric(0), source = character(0))
# Combine the vectors of CID into the data frame
q3_spr <- rbind(q3_spr, data.frame(value = df$MP_MOLvsTE_NormalizedSPR, source = rep("MP_MOLvsTE", length(df$MP_MOLvsTE_NormalizedSPR))))
q3_spr <- rbind(q3_spr, data.frame(value = df$ML_MOLvsTEASC_NormalizedSPR, source = rep("ML_MOLvsTEASC", length(df$ML_MOLvsTEASC_NormalizedSPR))))
q3_spr <- rbind(q3_spr, data.frame(value = df$ML_MOLvsTEnoASC_NormalizedSPR, source = rep("ML_MOLvsTEnoASC", length(df$ML_MOLvsTEnoASC_NormalizedSPR))))
```

Then, we need to statistically test whether MOL-TE incongruences (operationally measured as normalized RF, CID, and SPR) differ between (1) MP MOL vs TE, (2) ML MOL vs TE (with ASC), and (3) ML MOL vs TE (no ASC). Because data are not normally distributed (Shapiro-Wilk tests: P < 0.05) and more than one category is present in the predictor variable, we used Kruskal-Wallis tests for multiple categories and Mann-Whitney U tests for pairwise comparisons. 

- The Kruskal-Wallis test is a non-parametric test used to compare the medians (or ranks) of more than two groups. It is often used when you want to test whether there are statistically significant differences in the distribution of a continuous variable across multiple groups (i.e., a categorical factor with more than two levels), without assuming normality in the data. First, data is ranked (from the lowest to the highest value). Then, the statistic H (which involves calculating the sum of ranks for each group and comparing it to the overall mean rank) and the degrees of freedom are calculated. Finally, p-values can be estimated by comparing the statistic H and chi-squared distribution (if H is large, the p-value will be small).

- The Mann-Whitney U test is a non-parametric test, meaning it doesn't assume that the data follows a normal distribution. Instead of comparing means (like the t-test), the Mann-Whitney U test compares the ranks of the data values across two groups.

```{r}
# RF
shapiro.test(q3_rf$value) # p < 0.05, non-parametric
# Kruskal-Wallis: non-parametric test for multiple groups (non-paired values)
kruskal.test(q3_rf$value ~ q3_rf$source)
# Wilcoxon: pairwise comparisons for paired values
pairwise.wilcox.test(q3_rf$value, q3_rf$source, p.adjust.method = "bonferroni")
# Friedman: non-parametric test for paired values and multiple groups
q3_rf_fri = data.frame(df$MP_MOLvsTE_NormalizedRF, df$ML_MOLvsTEASC_NormalizedRF, df$ML_MOLvsTEnoASC_NormalizedRF)
friedman.test(as.matrix(q3_rf_fri))

# CID
shapiro.test(q3_cid$value) # p < 0.05, non-parametric
# Kruskal-Wallis: non-parametric test for multiple groups
kruskal.test(q3_cid$value ~ q3_cid$source)
# Wilcoxon: pairwise comparisons
pairwise.wilcox.test(q3_cid$value, q3_cid$source, p.adjust.method = "bonferroni")
# Friedman: non-parametric test for paired values and multiple groups
q3_cid_fri = data.frame(df$MP_MOLvsTE_NormalizedCID, df$ML_MOLvsTEASC_NormalizedCID, df$ML_MOLvsTEnoASC_NormalizedCID)
friedman.test(as.matrix(q3_cid_fri))
# CALCULATE NEMENYI TESTS IN RCLOUD !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

# SPR
shapiro.test(q3_spr$value) # p < 0.05, non-parametric
# Kruskal-Wallis: non-parametric test for multiple groups
kruskal.test(q3_spr$value ~ q3_spr$source)
# Perform pairwise comparisons using the Wilcoxon test for each pair
pairwise.wilcox.test(q3_spr$value, q3_spr$source, p.adjust.method = "bonferroni")
# Friedman: non-parametric test for paired values and multiple groups
q3_spr_fri = data.frame(df$MP_MOLvsTE_NormalizedSPR, df$ML_MOLvsTEASC_NormalizedSPR, df$ML_MOLvsTEnoASC_NormalizedSPR)
friedman.test(as.matrix(q3_spr_fri))
```

In all cases, Kruskal-Wallis results indicated that there is no association between MOL-TE incongruences and optimality criteria. Pairwise Mann-Whitney U (Wilcoxon rank-sum) tests also indicated no differences of MOL-TE incongruences between each pair of analysis. Thus, the addition of phenomic evidence is equally impactful on ML and MP topologies.

## Q3.2 Number of politomies

One importance difference, however, is the fact that trees were always binary in ML MOL and ML TE, whereas the addition of phenomic evidence can maintain, increase or decrease resolution of MP trees.

```{r}
# MP: Number of datasets in which the number of politomies is higher in TE than MOL
sum((df$MP_nPolytomiesTE - df$MP_nPolytomiesMOL) > 0)

# MP: Number of datasets in which the number of politomies is higher in MOL than TE
sum((df$MP_nPolytomiesTE - df$MP_nPolytomiesMOL) < 0)

# MP: Number of datasets in which the number of politomies is equal in MOL and TE
sum((df$MP_nPolytomiesTE - df$MP_nPolytomiesMOL) == 0)
```

## Q3.3 Contribution of phenomic evidence to the TE optimality score

In addition to testing whether MOL-TE incongruences differ among optimaliy criteria, we can explore the impact of phenomic evidence to optimality scores between optimality criteria. Specifically, it is possible that the MKv model can constrain the contribution of the phenomic partition to the TE optimality score in ML, whereas the same does not occur in MP analyses. As such, we need to test whether the proportion of optimality scores between MORPH and TE is higher in MP than ML analyses. 

```{r}
# Load data
df <- read.csv("Dataset_R_q123_06-03-25.csv")

# Exploratory data analyses
sum(df$MP_MORPH.TE_Score > df$ML_MORPH.TeASC_Score) # number of times that %MORPH/TE is higher in MP than ML ASC
sum(df$MP_MORPH.TE_Score < df$ML_MORPH.TeASC_Score) # number of times that %MORPH/TE is lower in MP than ML ASC

mean(df$MP_MORPH.TE_Score)
sd(df$MP_MORPH.TE_Score)
min(df$MP_MORPH.TE_Score)
max(df$MP_MORPH.TE_Score)

mean(df$ML_MORPH.TeASC_Score)
sd(df$ML_MORPH.TeASC_Score)
min(df$ML_MORPH.TeASC_Score)
max(df$ML_MORPH.TeASC_Score)

mean(df$ML_MORPH.TEnoASC_Score)
sd(df$ML_MORPH.TEnoASC_Score)
min(df$ML_MORPH.TEnoASC_Score)
max(df$ML_MORPH.TEnoASC_Score)

# Normality tests
shapiro.test(df$MP_MORPH.TE_Score) # p < 0.05, non-parametric
shapiro.test(df$ML_MORPH.TeASC_Score) # p < 0.05, non-parametric
shapiro.test(df$ML_MORPH.TEnoASC_Score) # p < 0.05, non-parametric

# Kruskal-Wallis (non-parametric, multiple groups, non-paired values)
y = c(df$MP_MORPH.TE_Score, df$ML_MORPH.TeASC_Score, df$ML_MORPH.TEnoASC_Score)
x = factor(rep(c("MP MOL vs MP TE", "ML MOL vs ML TE ASC", "ML MOL vs ML TE noASC"), each = length(df$MP_MORPH.TE_Score)))
kruskal.test(y ~ x)

# Friedman test (non-parametric, multiple groups, paired values)
q3_fri = data.frame(df$MP_MORPH.TE_Score, df$ML_MORPH.TeASC_Score, df$ML_MORPH.TEnoASC_Score)
friedman.test(as.matrix(q3_fri))

# Nemenyi test (post-hoc pairwise test, non-parametric, paired values). Run in RCLOUD
values <- c(df$MP_MORPH.TE_Score, df$ML_MORPH.TeASC_Score, df$ML_MORPH.TEnoASC_Score)
group <- c(rep("MP", length(df$MP_MORPH.TE_Score)),
             rep("ML ASC", length(df$ML_MORPH.TeASC_Score)),
             rep("ML noASC", length(df$ML_MORPH.TEnoASC_Score)))
block = rep(1:length(df$MP_MORPH.TE_Score), 3)  # Indices (same length as vectors)
df_nemenyi <- data.frame(value = values, group = group, block=block)
#PMCMRplus::frdAllPairsNemenyiTest(y = df_nemenyi$value, 
#                                  groups = df_nemenyi$group,
#                                  blocks = df_nemenyi$block)

# Figure
df_nemenyi$group <- factor(df_nemenyi$group, levels = c("MP", "ML ASC", "ML noASC"))
custom_palette <- c("MP" = "#339999", "ML ASC" = "#6f77b4", "ML noASC" = "#cccccc")

ggplot(df_nemenyi, aes(x = group, y = value, group = block)) +
  #geom_boxplot(outlier.shape = NA) +  # Create boxplot without default outliers
  geom_line(aes(group = block), color = "black", alpha = 0.1) +  # Link paired values
  geom_point(aes(color = group), alpha = .4, width = 0.2) +  # Add transparent dots
  labs(x = "", y = "% Phenomic/TE optimality score\n") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), legend.position = "bottom", legend.title = element_blank()) +  # Hide legend if not needed
  scale_color_manual(values = custom_palette)
```

## Q3.4 Figures

```{r}
# Set order of comparisons
q3_rf$source <- factor(q3_rf$source, levels = c("MP_MOLvsTE", "ML_MOLvsTEASC", "ML_MOLvsTEnoASC"))
q3_cid$source <- factor(q3_cid$source, levels = c("MP_MOLvsTE", "ML_MOLvsTEASC", "ML_MOLvsTEnoASC"))
q3_spr$source <- factor(q3_spr$source, levels = c("MP_MOLvsTE", "ML_MOLvsTEASC", "ML_MOLvsTEnoASC"))
# Set colors
custom_palette <- c("MP_MOLvsTE" = "#339999", "ML_MOLvsTEASC" = "#6f77b4", "ML_MOLvsTEnoASC" = "#cccccc")
# Create custom legend labels
custom_labels <- c(
  "MP_MOLvsTE" = "MP-MOL vs MP-TE",
  "ML_MOLvsTEASC" = "ML-MOL vs ML-TE (with ASC)",
  "ML_MOLvsTEnoASC" = "ML-MOL vs ML-TE (without ASC)"
)

# Boxplot Normalized RF
plot_q3_rf = ggplot(q3_rf, aes(x = source, y = value)) +
  geom_boxplot(outlier.shape = NA) +  # Create boxplot without default outliers
  geom_jitter(aes(color = source), alpha = .4, width = 0.2) +  # Add transparent dots
  labs(x = "", y = "Normalized RF\n", tag="A") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), legend.position = "none") +  # Hide legend if not needed
  scale_color_manual(values = custom_palette, labels = custom_labels)
# Boxplot CID
plot_q3_cid = ggplot(q3_cid, aes(x = source, y = value)) +
  geom_boxplot(outlier.shape = NA) +  # Create boxplot without default outliers
  geom_jitter(aes(color = source), alpha = .4, width = 0.2) +  # Add transparent dots
  labs(x = "", y = "\nNormalized CID\n", tag="B") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),legend.position = "none") +  # Hide legend if not needed
  scale_color_manual(values = custom_palette, labels = custom_labels)
# Boxplot SPR
plot_q3_spr = ggplot(q3_spr, aes(x = source, y = value)) +
  geom_boxplot(outlier.shape = NA) +  # Create boxplot without default outliers
  geom_jitter(aes(color = source), alpha = .4, width = 0.2) +  # Add transparent dots
  labs(x = "", y = "\nSPR moves\n", tag="C") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank())+
#theme(axis.text.x = element_blank(),legend.position = "none") +  # Hide legend if not needed
  scale_color_manual(values = custom_palette, labels = custom_labels)

# Plot 3x1
#tiff("Figures/figQ3_boxplots.tiff", width = 3, height = 6, units = "in", res = 300)
grid.arrange(plot_q3_rf + theme(plot.margin = margin(.01,1,1,.01)),
             plot_q3_cid + theme(plot.margin = margin(.001,1,1,.01)),
             plot_q3_spr + theme(plot.margin = margin(.001,1,1,.01)),
             nrow=3, ncol=1)
#dev.off()
```


# Q4. How do bootstrap values of shared clades differ between MOL and TE trees? 

First, we need to load the data. Support values of shared clades were identified and organized using the package RNODE. See script 2.

```{r}
# Load data
df_q4 = read.csv("Dataset_R_q4_06-03-25.csv")

# Set order of optimality criteria
df_q4$Optimality <- factor(df_q4$Optimality, levels = c("MP", "ML-ASC", "ML-noASC"))
df_q4$SupportDiff = df_q4$Support_Tree_1 - df_q4$Support_Tree_2
```

For each optimality criterion, we tested whether the paired values of bootstrap of shared clades differ between MOL and TE trees. First, we computed the number of clades in which bootstrap values were the same in MOL and TE trees, higher in MOL than TE trees, or higher in TE than MOL trees. Next, we tested the normality of bootstrap values using Shapiro-Wilk tests. Finally, we used both the non-parametric Wilcoxon signed-rank and Generalized Linear Mixed Models with Binomial distribution.

## Q4.1 MP MOL vs MP TE

```{r}
# Filter dataset keeping only MP
df_MP <- df_q4[grepl("MP", df_q4$Optimality), ]

# Exploratory data analysis
mean(df_MP$Support_Tree_1) # MOL
sd(df_MP$Support_Tree_1) # MOL

mean(df_MP$Support_Tree_2) # TE
sd(df_MP$Support_Tree_2) # TE

nrow(df_MP)# total number of shared clades between MOL and TE trees
sum(df_MP$Support_Tree_1 == df_MP$Support_Tree_2) # number of clades with BS MOL = BS TE
sum(df_MP$Support_Tree_1 > df_MP$Support_Tree_2) # number of clades with BS MOL > BS TE
sum(df_MP$Support_Tree_1 < df_MP$Support_Tree_2) # number of clades with BS MOL < BS TE

hist(1/df_MP$Support_Tree_1, breaks = 40, col = "lightblue", xlab = "Support Values")
hist(1/df_MP$Support_Tree_2, breaks = 40, col = "lightblue", xlab = "Support Values")
```

Data exploration of MP analyses revealed 43.4% (806/1858) of the clades whose bootstrap values were the same between MOL and TE analyses, 20.4% (380/1858) of the clades whose bootstrap values were higher in MOL than TE analyses, and 36.2% (672/1858) of the clades whose bootstrap values were higher in TE than MOL analyses. Moreover, the bootstrap in MOL clades was 89.2 ± 14.5, whereas the boostrap in TE clades was 90.4 ± 13.8.

```{r}
# Shapiro-Wilk test for normality
shapiro.test(df_MP$Support_Tree_1) # P < 0.05, non-parametric
shapiro.test(df_MP$Support_Tree_2) # P < 0.05, non-parametric

# Perform the Wilcoxon signed-rank test for paired values
wilcox.test(as.numeric(df_MP$Support_Tree_1),
                           as.numeric(df_MP$Support_Tree_2),
                           paired = TRUE)

# Dataset for GLMMs
df_MP_glmm <- data.frame(
  Node = c(df_MP$Node_Tree_1, df_MP$Node_Tree_2),
  Approach = c(df_MP$Approach1, df_MP$Approach2),
  Support = c(df_MP$Support_Tree_1, df_MP$Support_Tree_2),
  ID = c(df_MP$ID, df_MP$ID)
)

# GLMM Binomial considering the paired structure of shared clades
df_MP_glmm$Successes <- df_MP_glmm$Support
df_MP_glmm$Failures <- 100 - df_MP_glmm$Support
mp_binomial <- glmmTMB(cbind(Successes, Failures) ~ Approach + (1 | ID),
                       family = binomial(link = "logit"),
                       zi = ~1,
                       data = df_MP_glmm)
summary(mp_binomial)
```

## Q4.2 ML MOL vs ML TE ASC

```{r}
# Filter dataset keeping only MP
df_MLASC <- df_q4[grepl("ML-ASC", df_q4$Optimality), ]

# Exploratory data analysis
mean(df_MLASC$Support_Tree_1) # MOL
sd(df_MLASC$Support_Tree_1) # MOL

mean(df_MLASC$Support_Tree_2) # TE
sd(df_MLASC$Support_Tree_2) # TE

nrow(df_MLASC)# total number of shared clades between MOL and TE trees
sum(df_MLASC$Support_Tree_1 == df_MLASC$Support_Tree_2) # number of clades with BS MOL = BS TE
sum(df_MLASC$Support_Tree_1 > df_MLASC$Support_Tree_2) # number of clades with BS MOL > BS TE
sum(df_MLASC$Support_Tree_1 < df_MLASC$Support_Tree_2) # number of clades with BS MOL < BS TE

hist(1/df_MLASC$Support_Tree_1, breaks = 40, col = "lightblue", xlab = "Support Values")
hist(1/df_MLASC$Support_Tree_2, breaks = 40, col = "lightblue", xlab = "Support Values")
```

Data exploration of ML (with TE ASC) analyses revealed 47.1% (1089/2315) of the clades whose bootstrap values were the same between MOL and TE analyses, 26.6% (616/2315) of the clades whose bootstrap values were higher in MOL than TE analyses, and 26.3% (610/2315) of the clades whose bootstrap values were higher in TE than MOL analyses. Moreover, the bootstrap in MOL clades was 87.2 ± 19.1, whereas the boostrap in TE clades was 87.2 ± 19.6.

```{r}
# Shapiro-Wilk test for normality
shapiro.test(df_MLASC$Support_Tree_1) # P < 0.05, non-parametric
shapiro.test(df_MLASC$Support_Tree_2) # P < 0.05, non-parametric

# Perform the Wilcoxon signed-rank test for paired values
wilcox.test(as.numeric(df_MLASC$Support_Tree_1),
                       as.numeric(df_MLASC$Support_Tree_2),
                       paired = TRUE)

# Dataset for GLMMs
df_MLASC_glmm <- data.frame(
  Node = c(df_MLASC$Node_Tree_1, df_MLASC$Node_Tree_2),
  Approach = c(df_MLASC$Approach1, df_MLASC$Approach2),
  Support = c(df_MLASC$Support_Tree_1, df_MLASC$Support_Tree_2),
  ID = c(df_MLASC$ID, df_MLASC$ID)
)

# GLMM Binomial considering the paired structure of shared clades
df_MLASC_glmm$Successes <- df_MLASC_glmm$Support
df_MLASC_glmm$Failures <- 100 - df_MLASC_glmm$Support
mlasc_binomial <- glmmTMB(cbind(Successes, Failures) ~ Approach + (1 | ID),
                       family = binomial(link = "logit"),
                       zi = ~1,
                       data = df_MLASC_glmm)
summary(mlasc_binomial)
```

## Q4.3 ML MOL vs ML TE noASC

```{r}
# Filter dataset keeping only MP
df_MLnoASC <- df_q4[grepl("ML-noASC", df_q4$Optimality), ]

# Exploratory data analysis
mean(df_MLnoASC$Support_Tree_1) # MOL
sd(df_MLnoASC$Support_Tree_1) # MOL

mean(df_MLnoASC$Support_Tree_2) # TE
sd(df_MLnoASC$Support_Tree_2) # TE

nrow(df_MLnoASC)# total number of shared clades between MOL and TE trees
sum(df_MLnoASC$Support_Tree_1 == df_MLnoASC$Support_Tree_2) # number of clades with BS MOL = BS TE
sum(df_MLnoASC$Support_Tree_1 > df_MLnoASC$Support_Tree_2) # number of clades with BS MOL > BS TE
sum(df_MLnoASC$Support_Tree_1 < df_MLnoASC$Support_Tree_2) # number of clades with BS MOL < BS TE

hist(1/df_MLnoASC$Support_Tree_1, breaks = 40, col = "lightblue", xlab = "Support Values")
hist(1/df_MLnoASC$Support_Tree_2, breaks = 40, col = "lightblue", xlab = "Support Values")
```

```{r}
# Shapiro-Wilk test for normality
shapiro.test(df_MLnoASC$Support_Tree_1) # P < 0.05, non-parametric
shapiro.test(df_MLnoASC$Support_Tree_2) # P < 0.05, non-parametric

# Perform the Wilcoxon signed-rank test for paired values
wilcox.test(as.numeric(df_MLnoASC$Support_Tree_1),
                       as.numeric(df_MLnoASC$Support_Tree_2),
                       paired = TRUE)

# Dataset for GLMMs
df_MLnoASC_glmm <- data.frame(
  Node = c(df_MLnoASC$Node_Tree_1, df_MLnoASC$Node_Tree_2),
  Approach = c(df_MLnoASC$Approach1, df_MLnoASC$Approach2),
  Support = c(df_MLnoASC$Support_Tree_1, df_MLnoASC$Support_Tree_2),
  ID = c(df_MLnoASC$ID, df_MLnoASC$ID)
)

# GLMM Binomial considering the paired structure of shared clades
df_MLnoASC_glmm$Successes <- df_MLnoASC_glmm$Support
df_MLnoASC_glmm$Failures <- 100 - df_MLnoASC_glmm$Support
mlnoasc_binomial <- glmmTMB(cbind(Successes, Failures) ~ Approach + (1 | ID),
                       family = binomial(link = "logit"),
                       zi = ~1,
                       data = df_MLnoASC_glmm)
summary(mlnoasc_binomial)
```

## Q4.4 Figure

Remove the # after the chunk name.

```{r}#
# Create a new column with the optimality criterion
df_MP_glmm$comparison = "MPmol_MPte"
df_MLASC_glmm$comparison = "MLmol_MLteAsc"
df_MLnoASC_glmm$comparison = "MLmol_MLteNoAsc"

# Merge all comparisons into the same dataset
df_plot = bind_rows(df_MP_glmm, df_MLASC_glmm, df_MLnoASC_glmm)
View(df_plot)

# Set order of comparisons
df_plot$comparison <- factor(df_plot$comparison, levels = c("MPmol_MPte", "MLmol_MLteAsc", "MLmol_MLteNoAsc"))
df_plot = df_plot[!is.na(df_plot$comparison), ]

# Set colors
custom_palette <- c("MPmol_MPte" = "#339999", "MLmol_MLteAsc" = "#6f77b4", "MLmol_MLteNoAsc" = "#cccccc")
# Create custom legend labels
custom_labels <- c(
  "MPmol_MPte" = "MP-MOL vs MP-TE",
  "MLmol_MLteAsc" = "ML-MOL vs ML-TE (with ASC)",
  "MLmol_MLteNoAsc" = "ML-MOL vs ML-TE (without ASC)"
)

# Box Plot MP
plot_mp = ggplot(df_MP_glmm, aes(x = Approach, y = Support)) +
  geom_boxplot(outlier.shape = NA) +  # Create boxplot without default outliers
  geom_jitter(color = "#339999", alpha = .05, width = 0.2) +  # Add transparent dots
  labs(x = "", y = "\nBootstrap\n", tag="A") +
  theme_minimal() +
  #theme(axis.text.x = element_blank(), legend.position = "none") +  # Hide legend if not needed
  scale_color_manual(values = custom_palette, labels = custom_labels)
# Box Plot ML ASC
plot_ml_asc = ggplot(df_MLASC_glmm, aes(x = Approach, y = Support)) +
  geom_boxplot(outlier.shape = NA) +  # Create boxplot without default outliers
  geom_jitter(color = "#6f77b4", alpha = .05, width = 0.2) +  # Add transparent dots
  labs(x = "", y = "\nBootstrap\n", tag="B") +
  theme_minimal() +
  scale_y_continuous(limits = c(50, NA)) + # Define o mínimo para 50, o máximo será ajustado automaticamente
  #theme(axis.text.x = element_blank(), legend.position = "none") +  # Hide legend if not needed
  scale_color_manual(values = custom_palette, labels = custom_labels)
# Box Plot ML noASC
plot_ml_noasc = ggplot(df_MLnoASC_glmm, aes(x = Approach, y = Support)) +
  geom_boxplot(outlier.shape = NA) +  # Create boxplot without default outliers
  geom_jitter(color = "#cccccc", alpha = .08, width = 0.2) +  # Add transparent dots
  labs(x = "", y = "\nBootstrap\n", tag="C") +
  theme_minimal() +
  scale_y_continuous(limits = c(50, NA)) + # Define o mínimo para 50, o máximo será ajustado automaticamente
  #theme(axis.text.x = element_blank(), legend.position = "none") +  # Hide legend if not needed
  scale_color_manual(values = custom_palette, labels = custom_labels)

# Extract the legend as a separate ggplot object
legend <- ggplot(df_plot, aes(x = comparison, y = Support, color = comparison)) +
  geom_point() +
  scale_color_manual(values = custom_palette, labels = custom_labels) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid = element_blank(),
    panel.background = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
  )+
  guides(color = guide_legend(override.aes = list(size = 2, alpha=.7)))  # Increase dot size in legend
# Convert the plot to a grob
plot_grob <- ggplotGrob(legend)
# Extract the legend from the grob
legend_grob <- gtable::gtable_filter(plot_grob, "guide-box")

# Plot 2x2
#tiff("../../GitHub/Figures/figQ4_BS.tiff", width = 6, height = 6, units = "in", res = 300)
grid.arrange(arrangeGrob(plot_mp + theme(plot.margin = margin(1,1,1,.01)),
                         plot_ml_asc + theme(plot.margin = margin(1,1,1,.01)),
                         plot_ml_noasc + theme(plot.margin = margin(1,1,1,.01)),
                         nrow=1, ncol=3),
             legend_grob,
             nrow=2,
             heights= c(20,1))
#dev.off()
```

# Q5. Can bootstrap from MOL clades retrodict the monophyly of TE groups?

The dataset used in the Q5 was prepared with the package RNODE using the script 3.

```{r}
# Load data
df_q5 = read.csv("/Users/labanfibios/Desktop/Doutorado/Project/B2_TEvsMOL/GitHub/Dataset_R_q5_02-06-25.csv")

# Select MP 
MPlogistic <- df_q5[df_q5$optimality == "MP", ]
# Select ML ASC
MLASClogistic <- df_q5[df_q5$optimality == "MLASC", ]
# Select ML no ASC
MLnoASClogistic <- df_q5[df_q5$optimality == "MLnoASC", ]

```

## Q5.1 Data exploration

```{r}
#############################
# EXPLORATORY DATA ANALYSES #
#############################

# A) MP
# Total number of MOL clades
nrow(MPlogistic)
# Shared clades
nrow(subset(MPlogistic, occurrence_tree2 == 1)) # Number of shared clades
nrow(subset(MPlogistic, occurrence_tree2 == 1))/nrow(MPlogistic) # Proportion shared/total
mean(subset(MPlogistic, occurrence_tree2 == 1)$support_tree1) # BS mean
sd(subset(MPlogistic, occurrence_tree2 == 1)$support_tree1) # BS sd
range(subset(MPlogistic, occurrence_tree2 == 1)$support_tree1) # BS range
# Unique clades for MOL
nrow(subset(MPlogistic, occurrence_tree2 == 0)) # Number of unique clades for MOL
nrow(subset(MPlogistic, occurrence_tree2 == 0))/nrow(MPlogistic) # Proportion uniqueMOL/total
mean(subset(MPlogistic, occurrence_tree2 == 0)$support_tree1) # BS mean
sd(subset(MPlogistic, occurrence_tree2 == 0)$support_tree1) # BS sd
range(subset(MPlogistic, occurrence_tree2 == 0)$support_tree1) # BS range
# MOL clades with BS <75 present in TE
nrow(subset(MPlogistic, support_tree1 < 75 & occurrence_tree2 == 1)) # Number of MOL clades with BS <75 that occur in TE
nrow(subset(MPlogistic, support_tree1 < 75 & occurrence_tree2 == 1))/nrow(MPlogistic) # Proportion of MOL clades with BS <75 that occur in TE
# MOL clades with BS >90 absent in TE
nrow(subset(MPlogistic, support_tree1 > 90 & occurrence_tree2 == 0)) # Number of MOL lades with BS >90 absent in TE
nrow(subset(MPlogistic, support_tree1 > 90 & occurrence_tree2 == 0))/nrow(MPlogistic) # Proportion of MOL lades with BS >90 absent in TE
nrow(subset(MPlogistic, support_tree1 == 100 & occurrence_tree2 == 0)) # MOL clades with BS = 100 absent in TE

# B) ML ASC
# Total number of MOL clades
nrow(MLASClogistic)
# Shared clades
nrow(subset(MLASClogistic, occurrence_tree2 == 1)) # Number of shared clades
nrow(subset(MLASClogistic, occurrence_tree2 == 1))/nrow(MLASClogistic) # Proportion shared/total
mean(subset(MLASClogistic, occurrence_tree2 == 1)$support_tree1) # BS mean
sd(subset(MLASClogistic, occurrence_tree2 == 1)$support_tree1) # BS sd
range(subset(MLASClogistic, occurrence_tree2 == 1)$support_tree1) # BS range
# Unique clades for MOL
nrow(subset(MLASClogistic, occurrence_tree2 == 0)) # Number of unique clades for MOL
nrow(subset(MLASClogistic, occurrence_tree2 == 0))/nrow(MLASClogistic) # Proportion uniqueMOL/total
mean(subset(MLASClogistic, occurrence_tree2 == 0)$support_tree1) # BS mean
sd(subset(MLASClogistic, occurrence_tree2 == 0)$support_tree1) # BS sd
range(subset(MLASClogistic, occurrence_tree2 == 0)$support_tree1) # BS range
# MOL clades with BS <75 present in TE
nrow(subset(MLASClogistic, support_tree1 < 75 & occurrence_tree2 == 1)) # Number of MOL clades with BS <75 that occur in TE
nrow(subset(MLASClogistic, support_tree1 < 75 & occurrence_tree2 == 1))/nrow(MLASClogistic) # Proportion of MOL clades with BS <75 that occur in TE
# MOL clades with BS >90 absent in TE
nrow(subset(MLASClogistic, support_tree1 > 90 & occurrence_tree2 == 0)) # Number of MOL lades with BS >90 absent in TE
nrow(subset(MLASClogistic, support_tree1 > 90 & occurrence_tree2 == 0))/nrow(MLASClogistic) # Proportion of MOL lades with BS >90 absent in TE
nrow(subset(MLASClogistic, support_tree1 == 100 & occurrence_tree2 == 0)) # MOL clades with BS = 100 absent in TE


# C) ML noASC
# Total number of MOL clades
nrow(MLnoASClogistic)
# Shared clades
nrow(subset(MLnoASClogistic, occurrence_tree2 == 1)) # Number of shared clades
nrow(subset(MLnoASClogistic, occurrence_tree2 == 1))/nrow(MLnoASClogistic) # Proportion shared/total
mean(subset(MLnoASClogistic, occurrence_tree2 == 1)$support_tree1) # BS mean
sd(subset(MLnoASClogistic, occurrence_tree2 == 1)$support_tree1) # BS sd
range(subset(MLnoASClogistic, occurrence_tree2 == 1)$support_tree1) # BS range
# Unique clades for MOL
nrow(subset(MLnoASClogistic, occurrence_tree2 == 0)) # Number of unique clades for MOL
nrow(subset(MLnoASClogistic, occurrence_tree2 == 0))/nrow(MLnoASClogistic) # Proportion uniqueMOL/total
mean(subset(MLnoASClogistic, occurrence_tree2 == 0)$support_tree1) # BS mean
sd(subset(MLnoASClogistic, occurrence_tree2 == 0)$support_tree1) # BS sd
range(subset(MLnoASClogistic, occurrence_tree2 == 0)$support_tree1) # BS range
# MOL clades with BS <75 present in TE
nrow(subset(MLnoASClogistic, support_tree1 < 75 & occurrence_tree2 == 1)) # Number of MOL clades with BS <75 that occur in TE
nrow(subset(MLnoASClogistic, support_tree1 < 75 & occurrence_tree2 == 1))/nrow(MLnoASClogistic) # Proportion of MOL clades with BS <75 that occur in TE
# MOL clades with BS >90 absent in TE
nrow(subset(MLnoASClogistic, support_tree1 > 90 & occurrence_tree2 == 0)) # Number of MOL lades with BS >90 absent in TE
nrow(subset(MLnoASClogistic, support_tree1 > 90 & occurrence_tree2 == 0))/nrow(MLnoASClogistic) # Proportion of MOL lades with BS >90 absent in TE
nrow(subset(MLnoASClogistic, support_tree1 == 100 & occurrence_tree2 == 0)) # MOL clades with BS = 100 absent in TE

```

## Q5.2 Logistic regressions

```{r}

#######################
# LOGISTIC REGRESSION #
#######################

# A) MP
# Perform logistic regression
mp_logistic_result <- glm(occurrence_tree2 ~ support_tree1, data = MPlogistic, family = binomial)
# Print a summary of the model
summary(mp_logistic_result)
# Calculate 95% confidence intervals for the coefficients
confint(mp_logistic_result)
# Interpretation
exp(mp_logistic_result$coefficients[[2]]) # If coefficient is 0.09 and exp(0.09) is 1.1, then each unit of 1% bootstrap in MOL clade increases the chance of the occurrence of this clade in TE by 10%

# Plot the logistic regression curve with ggplot2
ggplot(MPlogistic, aes(x = support_tree1, y = occurrence_tree2)) +
  geom_point(alpha = 0.6) +  # Plot the data points
  geom_smooth(method = "glm", method.args = list(family = "binomial"), 
              color = "#339999", size = 1.2) +  # Logistic regression curve
  labs(x = "Bootstrap of MOL clade", y = "Probability of occurrence in the TE tree") +
  theme_minimal()

# B) ML ASC
# Perform logistic regression
mlasc_logistic_result <- glm(occurrence_tree2 ~ support_tree1, data = MLASClogistic, family = binomial)
# Print a summary of the model
summary(mlasc_logistic_result)
# Calculate 95% confidence intervals for the coefficients
confint(mlasc_logistic_result)
# Interpretation
exp(mlasc_logistic_result$coefficients[[2]]) # If coefficient is 0.06 and exp(0.09) is 1.06, then the each unit of 1% bootstrap in MOL clade increases the chance of the occurrence of this clade in TE by 6%

# Plot the logistic regression curve with ggplot2
ggplot(MLASClogistic, aes(x = support_tree1, y = occurrence_tree2)) +
  geom_point(alpha = 0.6) +  # Plot the data points
  geom_smooth(method = "glm", method.args = list(family = "binomial"), 
              color = "#6f77b4", size = 1.2) +  # Logistic regression curve
  labs(x = "Bootstrap of MOL clade", y = "Probability of occurrence in the TE tree") +
  theme_minimal()

# C) ML noASC
# Perform logistic regression
mlnoasc_logistic_result <- glm(occurrence_tree2 ~ support_tree1, data = MLnoASClogistic, family = binomial)
# Print a summary of the model
summary(mlnoasc_logistic_result)
# Calculate 95% confidence intervals for the coefficients
confint(mlnoasc_logistic_result)
# Interpretation
exp(mlnoasc_logistic_result$coefficients[[2]]) # If coefficient is 0.06 and exp(0.09) is 1.06, then the each unit of 1% bootstrap in MOL clade increases the chance of the occurrence of this clade in TE by 6%

# Plot the logistic regression curve with ggplot2
ggplot(MLnoASClogistic, aes(x = support_tree1, y = occurrence_tree2)) +
  geom_point(alpha = 0.6) +  # Plot the data points
  geom_smooth(method = "glm", method.args = list(family = "binomial"), 
              color = "#cccccc", size = 1.2) +  # Logistic regression curve
  labs(x = "\nBootstrap of MOL clade", y = "Probability of occurrence in the TE tree\n") +
  theme_minimal()
```

